---
title: 'US Storm Analysis'
output: html_document
---

##Synopsis
Storm Data is an official publication of the National Oceanic and Atmospheric Administration (NOAA), which documents several types of severe weather events that result in fatalities and or property damage.
In this, we try to summarize the data based on the number of fatalities or property damage for each event type and
try to figure out the worst weather events in terms of fatalities or property damage wise.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries
```{r libraries}
#library(cacher)
library(dplyr)
library(ggplot2)
```

## Reading data
Data Processing
We read the data in R and it is ready to be processed. No clean up activity is performed on the data.
```{r read, cache=TRUE}
data <- read.csv("C:/Users/SamJacobJulian/Downloads/repdata%2Fdata%2FStormData.csv.bz2")
head(data[,1:5])
```

## Description
There are several type of severe weather events that result in fatalities and or property damage.
In this block of code, we try to summarize the data based on the number of fatalities for each event type.
Since there are so many events that don't have any fatalities, they have been removed and the top 10 most fatal 
events have been listed in descending order.
```{r max_fatalities}
data_evtype <- data %>% group_by(EVTYPE) %>% summarise(sum(FATALITIES))
length(unique(data$EVTYPE))
# Separating events with fatalities
fat_ind <- which(data_evtype$`sum(FATALITIES)`>0)
fat <- data_evtype[fat_ind,]
# Top 10 disastrous events in US history
fat_ord <- fat %>% arrange(desc((fat$`sum(FATALITIES)`)))
head(fat_ord,10)
```

## Description
There are several type of severe weather events that result in fatalities and or property damage.
In this block of code, we try to summarize the data based on the extent of property damage for each event type.
Since there are so many events that don't have any property damage, they have been removed and the top 10 most damaging events have been listed in descending order.

```{r max_damage}
data_damage <- data %>% group_by(EVTYPE) %>% summarise(sum(PROPDMG))
# Separating events with fatalities
dam_ind <- which(data_damage$`sum(PROPDMG)`>0)
dam <- data_damage[dam_ind,]
# Top 10 disastrous events in US history
dam_ord <- dam %>% arrange(desc((dam$`sum(PROPDMG)`)))
head(dam_ord,10)
```

## Results
The results clearly show us that tornadoes are the most severe weather type in US.
```{r results}
# Finding maximum fatalities
which.max(data_evtype$`sum(FATALITIES)`)
print(data_evtype[834,])
# Finding maximum damages
which.max(data_damage$`sum(PROPDMG)`)
print(data_damage[834,])
```

## Plots
The plots show that tornadoes have been the most severe weather type for both human fatalities and property damage.
```{r plots}
sub1 <- head(fat_ord,10)
g1 <- ggplot(sub1,aes(sub1$EVTYPE,sub1$`sum(FATALITIES)`))+geom_point(col=1:10,size=3)
g1 <- g1 + ggtitle("US storm data: Fatalities") + xlab("Event type") + ylab("Human Fatalities")
g1 <- g1 + theme(axis.text.x = element_text(angle=90,hjust=1))
g1

sub2 <- head(dam_ord,10)
g2 <- ggplot(sub2,aes(sub2$EVTYPE,sub2$`sum(PROPDMG)`))+geom_point(col=11:20,size=3)
g2 <- g2 + ggtitle("US storm data: Property damage wise") + xlab("Event type") + ylab("Damages")
g2 <- g2 + theme(axis.text.x = element_text(angle=90,hjust=1))
g2
```
